---
title: "MOFA2 analyses with H&E image features"
author: "Sehyun Oh, Eslam Abousamra"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    theme: yeti
    highlight-style: tango
    toc: true
    toc-depth: 3
    toc-floating: true
abstract: "TCGA-COAD multi-modal"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

# Setup
## Load libraries
```{r include=FALSE,results="hide",message=FALSE,warning=FALSE}
library(MOFA2)
library(SpatialExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(HistoImageR)
library(ggplot2)
library(TCGAutils)
library(curatedTCGAData)
```

## Get data
```{bash eval=FALSE}
## All TCGA-COAD Prov-GigaPath outputs
wget -r -np -nH -R index.html https://u24-cancer-genomics.seandavi.workers.dev/provgigapath/TCGA_COAD/

## All TCGA-READ Prov-GigaPath outputs
wget -r -np -nH -R index.html https://u24-cancer-genomics.seandavi.workers.dev/provgigapath/TCGA_READ/
```

# Load data 
## Load embeddings
```{r}
dataDir <- "~/Packages/HistoImageR/data/provgigapath"
coad_dir <- file.path(dataDir, "TCGA_COAD")
coad_fnames <- list.files(coad_dir) %>% 
    grep(".csv", ., value = TRUE) %>%
    file.path(coad_dir, .)

# read_dir <- file.path(dataDir, "TCGA_READ")
# read_fnames <- list.files(read_dir) %>% 
#     grep(".csv", ., value = TRUE) %>%
#     file.path(read_dir, .)
# 
# fnames <- c(coad_fnames, read_fnames)
fnames <- coad_fnames
```

```{r embedding_layer0}
tensor_matrix <- matrix(nrow = length(fnames), ncol = 768)

## Simplify the slide id to participant id
participant_ids <- gsub(".csv", "", fnames) %>% 
    strsplit(., "-") %>% 
    sapply(., function(x) paste(x[1:3], collapse = "-"))
rownames(tensor_matrix) <- participant_ids

## Construct embedding (layer-0) matrix
for (i in seq_along(fnames)) {
  x <- readr::read_csv(fnames[i]) # load a file
  
  if (dim(x)[1] != 0) {
    res <- importEmbedding(x, layer = 0)
  } else {
    res <- NA
  }
  
  tensor_matrix[i,] <- res
}
```

```{r embedding_layer1, eval=FALSE}
tensor_matrix_layer1 <- matrix(nrow = length(fnames), ncol = 768)

## Simplify the slide id to participant id
participant_ids <- gsub(".csv", "", fnames) %>% 
    strsplit(., "-") %>% 
    sapply(., function(x) paste(x[1:3], collapse = "-"))
rownames(tensor_matrix_layer1) <- participant_ids

## Construct embedding (layer-0) matrix
for (i in seq_along(fnames)) {
  x <- readr::read_csv(fnames[i]) # load a file
  
  if (dim(x)[1] != 0) {
    res <- importEmbedding(x, layer = 1)
  } else {
    res <- NA
  }
  
  tensor_matrix_layer1[i,] <- res
}
```

```{r echo=FALSE, eval=FALSE}
## Slides without Prov-GigaPath results
no_res <- tensor_matrix[!complete.cases(tensor_matrix), ] %>% rownames
no_res
```

We have `nrow(tensor_matrix)` embeddings from TCGA_COAD and TCGA_READ.
```{r embedding_matrix}
tensor_matrix_full <- na.omit(tensor_matrix)
```


# Combine multiple modalities
MOFA2 doesn't directly support RaggedExperiment objects as one of the experiment 
types within a MultiAssayExperiment. The `create_mofa_from_MultiAssayExperiment()`
function specifically looks for experiments that can be converted to matrices 
using the `assay()` function; i.e., objects that have a standard `assay()` method 
returning a matrix. This includes SummarizedExperiment, SingleCellExperiment,
and RangedSummarizedExperiment objects. 

```{r}
#------------------------------------------
# 1. Load COAD + READ RNASeq assays
#------------------------------------------
# # Load COAD
# coad_mae <- curatedTCGAData(
#   diseaseCode = "COAD",
#   assays = "RNASeq2GeneNorm",
#   version = "2.1.1",
#   dry.run = FALSE
# )
# 
# # Load READ
# read_mae <- curatedTCGAData(
#   diseaseCode = "READ",
#   assays = "RNASeq2GeneNorm",
#   version = "2.1.1",
#   dry.run = FALSE
# )
# 
# rna_coad <- coad_mae[["COAD_RNASeq2GeneNorm-20160128"]]
# rna_read <- read_mae[["READ_RNASeq2GeneNorm-20160128"]]

coad_mae_all <- curatedTCGAData(
  diseaseCode = "COAD",
  assays = "*",
  version = "2.1.1",
  dry.run = FALSE
)

assays_subset <- c(
    "COAD_GISTIC_AllByGene-20160128", # CNA
    "COAD_RNASeq2GeneNorm-20160128", # RNAseq
    "COAD_Methylation_methyl450-20160128") # Methylation
coad_mae <- coad_mae_all[,,which(names(coad_mae_all) %in% assays_subset)]

#------------------------------------------
# 2. Ensure consistent genes
#------------------------------------------
# common_genes <- intersect(rownames(rna_coad), rownames(rna_read))
# rna_coad_bulk <- rna_coad[common_genes, ]
# rna_read_bulk <- rna_read[common_genes, ]
# 
# # Merge RNASeq matrices
# rna_combined <- cbind(rna_coad_bulk, rna_read_bulk)

# Standardize patient IDs (first 3 TCGA fields)
updated_coad_mae <- vector(mode = "list", length = length(coad_mae))
names(updated_coad_mae) <- names(coad_mae)
for (i in seq_along(coad_mae)) {
    mae <- coad_mae[[i]]
    patients <- sapply(strsplit(colnames(mae), "-"), function(x) paste(x[1:3], collapse = "-"))
    colnames(mae) <- make.unique(patients)
    updated_coad_mae[[i]] <- mae
}


#------------------------------------------
# 3. Prepare embedding matrix
#------------------------------------------
## Layer 0
tensor_matrix_unique <- tensor_matrix_full[!duplicated(rownames(tensor_matrix_full)), ] # keep only one slide per patient
embedding_patients <- sapply(strsplit(rownames(tensor_matrix_unique), "-"), function(x) paste(x[1:3], collapse = "-"))
rownames(tensor_matrix_unique) <- make.unique(embedding_patients) %>% basename

## Layer 1
tensor_matrix_unique1 <- tensor_matrix_layer1[!duplicated(rownames(tensor_matrix_layer1)), ] # keep only one slide per patient
embedding_patients <- sapply(strsplit(rownames(tensor_matrix_unique1), "-"), function(x) paste(x[1:3], collapse = "-"))
rownames(tensor_matrix_unique1) <- make.unique(embedding_patients) %>% basename
```

Subset to clinical metadata attributes with >95% completeness and remove technical fields
```{r}
#------------------------------------------
# 4. Prepare clinical metadata
#------------------------------------------
# Subset COAD clinical metadata
coad_clinical <- colData(coad_mae)
completeness <- colSums(!is.na(coad_clinical)) / nrow(coad_clinical) * 100
clinical_sub <- coad_clinical[, which(completeness > 95)] %>%
    as.data.frame() %>%
    select(-matches("aliquot"), 
           -matches("portion"))

# # Subset READ clinical metadata
# read_clinical <- colData(read_mae)
# completeness <- colSums(!is.na(read_clinical)) / nrow(read_clinical) * 100
# read_clinical_sub <- read_clinical[, which(completeness > 95)] %>%
#     as.data.frame() %>%
#     select(-matches("aliquot"), 
#            -matches("portion"))

# # Ensure all columns exist in both datasets
# all_cols <- union(colnames(coad_clinical_sub), colnames(read_clinical_sub))

# # Fill missing columns with NA
# coad_clinical_full <- coad_clinical_sub
# for (col in setdiff(all_cols, colnames(coad_clinical_sub))) coad_clinical_full[[col]] <- NA
# 
# read_clinical_full <- read_clinical_sub
# for (col in setdiff(all_cols, colnames(read_clinical_sub))) read_clinical_full[[col]] <- NA
# 
# # Merge COAD + READ clinical metadata
# clinical_sub <- rbind(coad_clinical_full, read_clinical_full)
# 
# # Keep only patients with embeddings
# patients_with_embed <- rownames(tensor_matrix_unique) %>% basename(.)
# clinical_sub <- clinical_sub[rownames(clinical_sub) %in% patients_with_embed, ]

# Factorize categorical variables with 2–10 levels
numLv <- apply(clinical_sub, 2, function(x) length(unique(x)))
slideAttrNames <- names(numLv[numLv >= 2 & numLv <= 10])
for (slideAttrName in slideAttrNames) {
  clinical_sub[[slideAttrName]] <- as.factor(clinical_sub[[slideAttrName]])
}

metadata_df <- clinical_sub[slideAttrNames]

# #------------------------------------------
# # 5. Subset RNASeq to patients with embeddings
# #------------------------------------------
# rna_matched <- coad_mae[, colnames(coad_mae) %in% patients_with_embed, drop = FALSE]
# 
# # Remove genes with Inf/-Inf
# rna_matrix <- assay(rna_matched)
# rna_keep <- apply(rna_matrix, 1, function(x) all(is.finite(x)))
# rna_matched_clean <- rna_matched[rna_keep, ]
# 
# # Subset embedding to match RNA columns
# embedding_matched_clean <- tensor_matrix_unique[colnames(rna_matched_clean), , drop = FALSE]

#------------------------------------------
# 6. Build SummarizedExperiment objects
#------------------------------------------
# rna_se <- SummarizedExperiment(
#   assays = list(counts = assay(rna_matched_clean)),
#   colData = DataFrame(patientID = colnames(rna_matched_clean))
# )

patients_with_image <- intersect(rownames(metadata_df), rownames(tensor_matrix_unique))
embedding_se_coldata <- metadata_df[patients_with_image,]

## layer0
embedding_se_subset <- tensor_matrix_unique[patients_with_image,]
embedding_se <- SummarizedExperiment(
  assays = list(embedding = t(embedding_se_subset)),
  colData = embedding_se_coldata
)
rownames(embedding_se) <- paste0("dim_", seq_len(nrow(embedding_se)))

# ## layer1
# embedding1_se_subset <- tensor_matrix_unique1[patients_with_image,]
# embedding1_se <- SummarizedExperiment(
#   assays = list(embedding = t(embedding1_se_subset)),
#   colData = embedding_se_coldata
# )
# rownames(embedding1_se) <- paste0("dim_", seq_len(nrow(embedding1_se)))


#------------------------------------------
# 7. Build MultiAssayExperiment
#------------------------------------------
cna_formatted <- updated_coad_mae[[1]]
rownames(cna_formatted) <- rowData(cna_formatted)[[1]]

crc_mae <- MultiAssayExperiment(
    experiments = list(
        # CNA = cna_formatted,
        RNAseq = updated_coad_mae[[2]],
        Methylation = updated_coad_mae[[3]],
        Embedding_layer0 = embedding_se
    ),
    colData = metadata_df
)
```


# MOFA2 analysis
## Create MOFA object
```{r}
# Create MOFA object
mofa_object <- create_mofa_from_MultiAssayExperiment(crc_mae)

# Define model options
model_opts <- get_default_model_options(mofa_object)
model_opts$num_factors <- 15

# Define training options
train_opts <- get_default_training_options(mofa_object)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 42

# Prepare and run MOFA
mofa_object <- prepare_mofa(
  object = mofa_object,
  model_options = model_opts,
  training_options = train_opts
)
```

## Run MOFA
+ `mofa_model_crc_ver0.hdf5`
    - RNAseq
    - Embedding_layer0
+ `mofa_model_crc_ver1.hdf5`
    - RNAseq
    - Methylation
    - Embedding_layer0
    - Embedding_layer1
+ `mofa_model_crc_ver2.hdf5`
    - CNA: COAD_GISTIC_AllByGene-2016012
    - RNAseq: COAD_RNASeq2GeneNorm-20160128
    - Methylation: COAD_Methylation_methyl450-20160128
    - Embedding_layer0
+ `mofa_model_crc_ver3.hdf5`
    - CNA: COAD_GISTIC_AllByGene-2016012
    - RNAseq: COAD_RNASeq2GeneNorm-20160128
    - Embedding_layer0
    
+ `mofa_model_crc_ver4.hdf5`
    - RNAseq
    - Methylation
    - Embedding_layer0

```{r train-mofa-model}
mofa_object.trained <- run_mofa(
    mofa_object,
    outfile = "mofa_model_crc_ver4.hdf5",
    use_basilisk = TRUE
)
```

## Analyze MOFA model
```{r}
ver <- 4
    
# Load trained MOFA model
mofa_object.trained <- load_model(paste0("mofa_model_crc_ver", ver,".hdf5"))

# Keep only patients in MOFA object
mofa_samples <- unlist(samples_names(mofa_object.trained))
metadata_matched <- metadata_df[mofa_samples, , drop = FALSE]

# For the metadata data frame that we want to add to the MOFA object:
metadata_df$sample <- rownames(metadata_df)

# Now, when we set the samples_metadata, it will have the required 'sample' column.
samples_metadata(mofa_object.trained) <- metadata_df

# Save model with metadata
saveRDS(mofa_object.trained, paste0("mofa_model_crc_ver", ver,".rds"))
```
