---
title: "MOFA2 analyses with H&E image features"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    theme: yeti
    highlight-style: tango
    toc: true
    toc-depth: 3
    toc-floating: true
abstract: "Visualize MOFA models from TCGA-COAD multi-modal analysis"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```


# Setup
## Load libraries
```{r include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(MOFA2)
library(ggplot2)
library(dplyr)
```


# MOFA Visualizations
```{r}
ver <- 5
mofa_object.trained <- readRDS(paste0("MOFA_models/mofa_model_crc_ver", ver,".rds"))
```

```{r echo=FALSE, eval=FALSE}
ver <- 2
mofa_object.trained <- readRDS(paste0("MOFA_models/mofa_model_read_ver", ver,".rds"))
```

```{r reformat, echo=FALSE}
## Update long view names
rename_map <- data.frame(
    col1 = c("Methylation", 
             "Embedding_layer0"),
    col2 = c("DNAm", 
             "Embedding")
)

viewsNames <- views_names(mofa_object.trained)
views_names(mofa_object.trained) <- plyr::mapvalues(
    x = viewsNames, 
    from = rename_map$col1, 
    to = rename_map$col2, 
    warn_missing = FALSE)

## Update long column names
long_names_df <- data.frame(
    col1 = c("patient.microsatellite_instability_test_results.microsatellite_instability_test_result.mononucleotide_and_dinucleotide_marker_panel_analysis_status",
             "patient.samples.sample.2.sample_type",
             "patient.biospecimen_cqcf.top_slide_submitted",
             "gender.x",
             "vital_status.x",
             "histological_type.x",
             "patient.clinical_cqcf.frozen_specimen_anatomic_site"),
    col2 = c("MSI",
             "sample_type", 
             "top_slide",
             "gender",
             "vital_status",
             "histological_type",
             "specimen_site")
)

meta <- samples_metadata(mofa_object.trained)
colNames <- colnames(meta)
colnames(meta) <- plyr::mapvalues(
    x = colNames, 
    from = long_names_df$col1, 
    to = long_names_df$col2, 
    warn_missing = FALSE)
samples_metadata(mofa_object.trained) <- meta
```


```{r warning=FALSE, fig.width=4, fig.height=4, collapse = FALSE}
# Visualize the model
plot_data_overview(mofa_object.trained)

# Plot factor correlations
plot_factor_cor(mofa_object.trained)
```

```{r warning=FALSE, fig.width=4, fig.height=4, collapse = FALSE}
# Plot variance explained
plot_variance_explained(mofa_object.trained, plot_total = TRUE)
```

Choose the MOFA Factor number to check
```{r}
factor_num <- 1
```

```{r}
kw_res <- HistoImageR:::testFactorAssociations(mofa_object.trained, factor_num)
top_attrs <- kw_res %>% 
    filter(p_adjusted < 0.5) %>% 
    arrange(desc(statistic)) %>% 
    head
top_attrs
```

```{r warning=FALSE, fig.width=3.5, fig.height=4, collapse = FALSE}
# Plot factor colored by clinical data
clinical_attrs <- top_attrs$group

for (clinical_attr in clinical_attrs) {
    plt <- plot_factor(mofa_object.trained, 
            factor = factor_num, 
            add_violin = TRUE,
            color_by = clinical_attr,
            # legend = FALSE,
            dodge = TRUE)
    print(plt)
}
```

```{r warning=FALSE, fig.width=5, fig.height=4, collapse = FALSE, eval=FALSE, echo=FALSE}
# Plot factor scatter
plot_factor(mofa_object.trained, 
            factor = factor_num, 
            add_violin = TRUE,
            color_by = "patient.clinical_cqcf.frozen_specimen_anatomic_site",
            dodge = TRUE)

# # Plot factor colored by MYH11 gene
# plot_factor(mofa_object.trained, 
#             factor = 1, 
#             add_violin = TRUE,
#             dodge = TRUE,
#             color_by = "MYH11")
```

```{r warning=FALSE, fig.width=6, fig.height=4, collapse = FALSE}
## Plot data heatmaps
plot_data_heatmap(mofa_object.trained,
                  factor = factor_num,
                  view = "Embedding",
                  features = 100,
                  annotation_samples = c("sample_type"),
                  cluster_rows = TRUE,
                  cluster_cols = TRUE, 
                  show_colnames = FALSE,
                  show_rownames = FALSE)

samples_to_keep <- meta %>% filter(sample_type != "blood derived normal") %>% rownames(.)
trained_mofa_subset <- subset_samples(mofa_object.trained, samples = samples_to_keep)
plot_data_heatmap(trained_mofa_subset,
                  factor = factor_num,
                  view = "Embedding",
                  features = 100,
                  annotation_samples = c("sample_type", "top_slide", "pathology_N_stage"),
                  cluster_rows = TRUE,
                  cluster_cols = TRUE, 
                  show_colnames = FALSE,
                  show_rownames = FALSE)

# plot_data_heatmap(mofa_object.trained,
#                   factor = factor_num,
#                   view = "Embedding",
#                   features = 10,
#                   cluster_rows = TRUE,
#                   cluster_cols = FALSE,
#                   show_colnames = FALSE)
```

```{r warning=FALSE, fig.width=4, fig.height=4, collapse = FALSE, echo=FALSE, eval=FALSE}
## Plot top weights
plot_top_weights(mofa_object.trained,
                 view = "RNAseq",
                 factor = factor_num,
                 nfeatures = 10)

plot_top_weights(mofa_object.trained,
                 view = "Embedding", 
                 factor = factor_num,
                 nfeatures = 10)
```

A larger absolute feature weight indicates a stronger association of the 
feature with the factor. A sign interprets the effect direction: positive — the 
feature demonstrates larger values in the samples with positive factor values, etc.

```{r eval=FALSE}
## Need to debug
# Error in .check_and_get_views(object, view) : 
#   all(views %in% views_names(object)) is not TRUE
plot_weights(mofa_object.trained,
             view = "RNAseq",
             factor = factor_num,
             nfeatures = 5,
             scale = TRUE)
```

```{r eval=FALSE, echo=FALSE}
# Get top features for the selected Factor 
top_features <- get_weights(
    mofa_object.trained,
    factor = factor_num,
    view = "Protein", 
    abs = TRUE)
top_features_list <- top_genes[[1]][,1] %>% 
    sort(., decreasing = TRUE) %>% 
    names %>% .[1:10]
top_features_list
```

```{r enrichGO, echo=FALSE, eval=FALSE}
# Get top features for the selected Factor 
top_genes <- get_weights(
    mofa_object.trained,
    factor = factor_num,
    view = "RNAseq", 
    abs = TRUE)

top_genes_list <- top_genes[[1]][,1] %>% 
    sort(., decreasing = TRUE) %>% 
    names %>% .[1:100]

# Perform pathway enrichment
pathway_results <- clusterProfiler::enrichGO(top_genes_list, "org.Hs.eg.db")
```

```{r GSEA, echo=FALSE, eval=FALSE}
top_genes <- get_weights(
    mofa_object.trained,
    factor = factor_num,
    view = "RNAseq", 
    abs = FALSE)

top_genes_list <- top_genes[[1]][,1] %>% 
    sort(., decreasing = TRUE)


term2gene <- clusterProfiler::read.gmt("~/Packages/HistoImageR/inst/extdata/c2.cp.v7.1.symbols.gmt")
colnames(term2gene) <- c("gs_name", "entrez_gene")

gsea_result <- clusterProfiler::GSEA(
    top_genes_list, 
    TERM2GENE = term2gene,
    pvalueCutoff = 0.05, 
    seed = TRUE)

dotplot(gsea_result, 
        showCategory = 10,
        split = ".sign") + 
  facet_grid(.~.sign) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 5))

enrichplot::gseaplot2(gsea_result, 
          geneSetID = c(1, 2, 3),
          pvalue_table = TRUE,
          color = c("red", "blue", "green"),
          ES_geom = "line",
          subplots = 1:3)
```

Association analysis between the sample metadata and the Factor values
```{r}
# Factor association analysis
correlate_factors_with_covariates(mofa_object.trained, 
                                  covariates = c(
                                      # "gender","vital_status",
                                                 # "MSI",
                                                 "top_slide",
                                                 "histological_type",
                                                 # "pathology_T_stage", "pathology_N_stage", "pathology_M_stage",
                                                 "sample_type", "specimen_site"),
                                  # covariates = colnames(samples_metadata(mofa_object.trained)),
                                  plot = "r",
                                  factors = 1:10,
                                  cluster_cols = FALSE)
```

